% !TeX root = CategoricalSemantics.tex
\documentclass[../../Dissertation.tex]{subfiles}

\begin{document}
In the pursuit of a suitable categorical semantics for parametricity, a well-known isomorphism between formal proof systems, category theory and  programs, will be introduced. This will facilitate the exploration of a composable family of dinatural transformations, which are identified by means of cut-free proofs.
\par
A categorical semantics of the simply-typed lambda calculus will be provided.

\subsection{Curry-Howard-Lambek correspondence}\label{sec:curryhowardlambek}
A sequent calculus is an inference system for first-order logic. The most basic construct in a sequent calculus, known as a sequent, has the form
\begin{flalign}\label{eq:sequent}
  X_1,\ X_2,\ \ldots,\ X_N \vdash Y_1,\ Y_2,\ \ldots,\ Y_M.
\end{flalign}
A sequent of the form expressed in (\ref{eq:sequent}) is a judgement which asserts that a statement $Y_i$ is true, if $X_1 \land X_2 \land ... \land X_N$ is true. 
\par
Constructs in formal proof systems, such as a sequent calculus, are in direct correspondence with types within a typed lambda calculus, by the Curry-Howard isomorphism \cite{CurryHoward}. A sequent of the form
\begin{flalign}\label{eq:sequent}
  X_1,\ X_2,\ \ldots,\ X_N \vdash Y
\end{flalign}
corresponds to a term of the type $X_1 \times X_2 \times ... \times X_N \Rightarrow Y$, and constructing such a term is equivalent to the construction of a proof.
\par
Connections between intuitionistic logic and category theory, discovered by Lambek \citeyear{LambekCorrespondence,LambekCorrespondence2,LambekCorrespondence3} and \citeasnoun{LawvereCorrespondence}, further extends this relationship to a three-way isomorphism known as the Curry-Howard-Lambek correspondence. The Lambek-Lawvere correspondence can be described as modelling formulae in an intuitionistic propositional logic as objects in a cartesian-closed category, and proofs as morphisms.
\par
Let $\Gamma$, $\Sigma$, $\Delta$ and $\Lambda$ be finite lists of formulae, known as a context. The cut-rule is then defined as follows:

\begin{equation}\label{eq:cut}
\begin{prooftree}
  \hypo{\Gamma \vdash X, \Delta}
  \hypo{\Sigma, X \vdash \Lambda}
  \infer2[cut]{\Sigma, \Gamma \vdash \Lambda, \Delta}
\end{prooftree}.
\end{equation}
The cut-rule depicted in (\ref{eq:cut}) can be described as cutting any occurrence of $X$ from the proof. By the Curry-Howard isomorphism, the cut rule corresponds to a familiar substitution process; function composition. Given a context $\Pi$ in a typed lambda calculus, this substitution may be stated formally as
\begin{equation}\label{eq:lambdacompose}
\begin{prooftree}
  \hypo{\Pi, x : A \vdash t : B}
  \hypo{\Pi, y : B \vdash t' : C}
  \infer2[]{\Pi, x : A \vdash t'[t/y] : C}
\end{prooftree}.
\end{equation}
\par
The cut-elimination theorem, originally proved by \citeasnoun{CutFree}, states that any sequent which has a proof using the cut-rule, also has a proof which does not. Cut-free proofs are typically more complex and larger than alternative proofs which make use of the cut rule. Cut-elimination corresponds to the strong-normalization property of a typed lambda calculus.
\par
The substitution depicted in (\ref{eq:lambdacompose}) reveals a property of the cut-rule; covariant occurrences of proofs are substituted into contravariant occurrences. Conversely, the identity sequent maps contravariant occurrences of proofs to covariant occurrences. 
\par
\citeasnoun{NaturalAsNormalForms} describe how the symmetry induced by the co-existence of the identity sequent and the cut rule is broken by the cut-elimination theorem. This asymmetry is used to show that cut-elimination, when interpreted by dinatural transformations, resolves the cyclic dependencies created by `linking' covariant and contravariant terms. Using cut-free sequent rules, Girard, Scedrov and Scott show that syntactic dinatural transformations between definable functors on a cartesian-closed category, whose internal language is a typed lambda calculus, are composable.

\subsection{Categorical Semantics of System F}\label{sec:systemfcat}
A model of the simply-typed lambda calculus in a cartesian closed category was first described by \citeasnoun{CCCLambda}. Furthermore, Lambek shows the simply-typed lambda calculus is the internal language of any cartesian closed category. The notion of an internal language provides a manner in which to translate propositions within a given logic into categorical propositions. The construction described by Lambek extends to the second-order polymorphic lambda calculus (System F).
\par
Types in the simply-typed calculus are taken to be objects in a category. The set of morphisms between two types $T$ and $U$ is given by the set of functions of the form $T \Rightarrow U$. The identity morphism is given by the polymorphic identity function. Composition is given by the polymorphic composition function.
\par
To complete the construction of a cartesian closed category, it is necessary to show that the category has finite products and that the functor, $- \times X$, has a right adjoint. The terminal object is given by including a unit type, denoted $()$, in the definition of the simply-typed lambda  calculus. 
\par
Binary products are also included within the definition. For any types $T$ and $U$, there exists a type $T \times U$ and projections $\pi_{T} : T \times U \Rightarrow T$, $\pi_{U} : T \times U \Rightarrow U$, such that for any type $V$ and functions $f : V \Rightarrow T$, $g : V \Rightarrow U$, there exists a function $h : V \Rightarrow T \times U$, such that the following diagram commutes:

\begin{equation}\label{eq:nat_vertical}
  \begin{tikzpicture}[baseline= (a).base]
    \node[scale=1.2] (a) at (0,0) {
      \begin{tikzcd}
        & 
        V \arrow[dl, swap, "f"] \arrow[dr, "g"] \arrow[d, "h"]
        &
        \\
        T
        & 
        T \times U \arrow[l, "\pi_{T}"] \arrow[r, swap, "\pi_{U}"]
        &
        U
      \end{tikzcd}.
    };
  \end{tikzpicture}
\end{equation}
\\
The right adjoint of $- \times X$ is given by the functor $X \Rightarrow -$, which sends a type $Y$ to the function type $X \Rightarrow Y$, and a natural transformation $\epsilon_X : (X \Rightarrow Y) \times X \rightarrow Y$ given by evaluation. Given any function $f : Z \times X \Rightarrow Y$, there exists the unique function $g : Z \Rightarrow (X \Rightarrow Y) = z \mapsto (x \mapsto f (z, x))$, which makes the following diagram commute:
\begin{equation}\label{eq:nat_vertical}
  \begin{tikzpicture}[baseline= (a).base]
    \node[scale=1.2] (a) at (0,0) {
      \begin{tikzcd}
        (X \Rightarrow Y) \times X \arrow[r, "\epsilon_X"]
        &
        Y
        \\
        Z \times X \arrow[u, "g \times id"] \arrow[ur, swap, "f"]
        &
      \end{tikzcd}.
    };
  \end{tikzpicture}
\end{equation}

\subsection{Functorial Polymorphism}\label{sec:functorial}
\citeasnoun{FunctorialPolymorphism} introduce a formal model of parametric polymorphism in which the parametricity property is described by naturality conditions. Given an enriched cartesian closed category $\mathbb{C}$ of primitive types, multi-variance functors of the form $(\mathbb{C}^{op})^n \times \mathbb{C}^n \rightarrow \mathbb{C}$ are used to model polymorphic types with $n$ type variables. In this interpretation, terms are given by natural transformations and primitive types are modelled by constant functors.
\par
A deeper intuition of how types may be understood as functors of the form $(\mathbb{C}^{op})^n \times \mathbb{C}^n \rightarrow \mathbb{C}$, is found by examination of prominent examples. The type of the polymorphic identity function, $\forall T.\ T \Rightarrow T$, is modelled as a functor of the form $\mathbb{C}^{op} \times \mathbb{C} \rightarrow \mathbb{C}$. The type of the Y-combinator, $\forall T.\ (T \Rightarrow T) \Rightarrow T$, is modelled as a functor of the form $\mathbb{C}^{op} \times \mathbb{C}^{op} \times \mathbb{C} \times \mathbb{C} \rightarrow \mathbb{C}$, which is dummy in the first argument.
\par
Bainbridge et al. construct a cartesian closed category, $\mathbf{PER}(X)$, of partial-equivalence relations on $X$, in which a morphism is named by each partial recursive function $f$ which preserves relationships i.e., $x\ R\ y \rightarrow f(x)\ R'\ f(y)$. Composition is given by the set-theoretic notion of function composition. A class of composable dinatural transformations, denoted `realizable', on the category $\mathbf{PER}(X)$, is identified. It is then shown that any functor definable in System F is `realizable', and every natural transformation between such functors is indeed also `realizable'. Dinatural transformations within System F, defined using the described syntatic approach, are thus shown to be composable.
\par
\citeasnoun{DinaturalInSystemF} formulates a sufficient condition for dinaturality, in both the Church and Curry-style second-order polymorphic lambda calculus. Lataillade first describes the syntactic category $\mathcal{T}$ for System F, in conjunction with
functors of the form $F[(\_,\_),X] : \mathcal{T}^{op} \times \mathcal{T} \rightarrow \mathcal{T}$. Each functor $F[(A,B),X]$ represents a substitution in which contravariant occurrences of the type variable $X$ are replaced with $A$ and covariant occurrences are replaced with $B$. The dinaturality condition between these substitution functors are used to create a sufficient condition for dinaturality in System F. In the Church-style second-order polymorphic lambda calculus, a term $T$ is shown to be dinatural in a type variable $X$ if and only if $X$ is not in the set of free type variables across all type instantiations of the normal-form of $T$.

\subsection{Extensions to Functorial Polymorphism}
Discuss existing extensions to the functorial polymorphism and the `solutions' to the composition problem in the literature.
\newline\newline
\citeasnoun{StructuralPolymorphism} describe the notion of `structors'; a generalisation of functors which carry morphisms to spans of morphisms.
\newline\newline
\citeasnoun{CanonicalDinatural} define `canonical dinatural transformations', a constrained form of dinatural transformations, and relates them to context-free grammars and the polymorphic operator.

\end{document}
